---
title: "00 — Introduction & Setup"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

## What you're working with

This is a **fully simulated** case/control GWAS dataset — no real genomes
involved. It was built to behave like a real one, problems included:

- **10 fake chromosomes**, 5,000 SNPs each = **50,000 SNPs total**
- A **majority ancestry group** plus **two minority ancestry groups** mixed
  into the raw data, with no ancestry label given to you up front
- A handful of samples with **injected QC problems** (contamination-like
  excess heterozygosity, poor-quality high missingness)
- A block of SNPs with a genuine **Hardy-Weinberg violation** (simulated
  genotyping artefact) and a block with a **failing assay** (high
  missingness)
- Some **real disease-associated SNPs** baked in, at a range of frequencies
  and effect sizes — and one **decoy SNP** that will look associated purely
  because of population stratification, until you deal with it

You don't get to know which SNPs or samples are which — that's the point.
By the end of notebook 05 you'll have found (most of) it yourself, and you
can compare notes with the instructor's answer key afterward.

## One-time setup

If this is a fresh container, the dataset has already been generated for
you (via `postCreateCommand`). Confirm the files exist:

```{bash}
ls -la data/
```

If `data/` is empty, generate it now (takes well under a minute):

```{r generate-if-missing, eval=FALSE}
source("R/simulate_toy_gwas.R")
```

## Sanity-checking the data (without peeking at the answer key)

Before trusting a simulated — or any — dataset, check that it looks like
what it's supposed to look like. Run PLINK2's allele-frequency report:

```{bash}
mkdir -p results
plink2 --bfile data/study --freq --out results/sanity_freq
```

```{r}
freq <- data.table::fread("results/sanity_freq.afreq")
names(freq) <- sub("^#", "", names(freq))
head(freq)
```

A few checks worth doing on any new dataset before you analyze it:

```{r}
# How many variants and samples did we actually get?
bim <- data.table::fread("data/study.bim", header = FALSE)
fam <- data.table::fread("data/study.fam", header = FALSE)
cat("SNPs:   ", nrow(bim), "\n")
cat("Samples:", nrow(fam), "\n")
cat("Chromosomes present:", paste(sort(unique(bim$V1)), collapse = ", "), "\n")
```

```{r}
# Allele frequencies should mostly land in a sensible range for common SNPs
hist(pmin(freq$ALT_FREQS, 1 - freq$ALT_FREQS),
     breaks = 50, main = "Minor allele frequency, all SNPs", xlab = "MAF")
```

> **Note on ALT_FREQS vs. minor allele frequency:** PLINK's `ALT_FREQS`
> column is the frequency of *whichever allele happens to be labeled ALT*
> in the `.bim` file — not necessarily the minor allele. Notice above that
> we took `pmin(ALT_FREQS, 1 - ALT_FREQS)` to get the actual minor allele
> frequency. This trips people up in real GWAS output too: always check
> which allele a reported frequency (or later, an odds ratio) refers to.

**🔧 YOUR TURN:** the phenotype column in `data/study.fam` uses PLINK's
convention (`1` = control, `2` = case). Count how many of each are in the
raw file below — don't worry that the numbers look a little bigger/smaller
than "2,000 and 2,000"; you haven't done any QC yet, and this file
deliberately still contains some samples that shouldn't be here.

```{r}
table(fam$V6)  # ___FILL_IN___: is V6 really the phenotype column? Check ?read_fam or the .fam spec if unsure.
```

## What's next

`01_qc_missingness_heterozygosity.Rmd` starts the actual pipeline: sample
QC based on missingness and heterozygosity, where you'll pick your own
thresholds from the data rather than being handed one.
