---
title: "01 — Sample QC: Missingness & Heterozygosity"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
library(dplyr)
library(ggplot2)
```

## Why these two checks, together

Two of the most common signs of a bad DNA sample:

- **High missingness** — a poor-quality sample where many genotypes
  couldn't be called at all.
- **Abnormal heterozygosity** — usually too *high* (a classic sign of
  sample contamination: DNA from two people mixed together looks
  heterozygous almost everywhere), sometimes too *low* (inbreeding, or a
  labeling/technical error).

Plotted against each other, per sample, bad samples tend to stand apart
from the main cloud — and this is the standard first QC plot in real
case/control GWAS pipelines (Anderson et al. 2010 is the classic reference
protocol). Nobody hands you a fixed threshold in advance; you look at your
own data's distribution and decide.

## Step 1 — compute missingness and heterozygosity per sample

```{bash}
mkdir -p results
plink2 --bfile data/study --missing --out results/study_qc
plink2 --bfile data/study --het --out results/study_qc
```

`--missing` writes `results/study_qc.smiss` (per-sample) and `.vmiss`
(per-SNP, we'll come back to that one later). `--het` writes
`results/study_qc.het`.

## Step 2 — load both into R and inspect them

Don't assume column names — inspect the file before you write code against it:

```{r}
smiss <- data.table::fread("results/study_qc.smiss")
het   <- data.table::fread("results/study_qc.het")
names(smiss) <- sub("^#", "", names(smiss))
names(het)   <- sub("^#", "", names(het))
names(smiss)
names(het)
```

**🔧 YOUR TURN:** `--het` reports observed homozygote count, expected
homozygote count, and the number of non-missing genotypes it was computed
over. Using the column names printed above, compute each sample's
**observed heterozygosity rate**:

$$\text{het rate} = \frac{N_{\text{non-missing}} - O(\text{HOM})}{N_{\text{non-missing}}}$$

```{r}
het <- het %>%
  mutate(het_rate = (___FILL_IN___ - ___FILL_IN___) / ___FILL_IN___)
```

## Step 3 — the QC scatter plot

```{r}
qc <- smiss %>%
  select(FID, IID, F_MISS) %>%
  inner_join(het %>% select(FID, IID, het_rate), by = c("FID", "IID"))

het_mean <- mean(qc$het_rate)
het_sd   <- sd(qc$het_rate)

ggplot(qc, aes(x = F_MISS, y = het_rate)) +
  geom_point(alpha = 0.4, size = 1) +
  scale_x_log10() +
  labs(x = "Missingness rate (log scale)", y = "Heterozygosity rate",
       title = "Per-sample missingness vs. heterozygosity") +
  theme_minimal(base_size = 12)
```

Look at the shape of that cloud. Is there a dense main group and a
handful of points clearly off to the side? In which direction(s)?

## Step 4 — 🔧 YOUR TURN: pick your own thresholds

There's no single universally "correct" cutoff — you're making a judgment
call from the distribution you just plotted, the same way you would with
real data. A commonly used starting point (not a law of nature): flag
heterozygosity beyond **mean ± 3 SD**, and missingness above some rate that
clearly separates the main cloud from the outliers in *your* plot.

```{r}
het_threshold_low  <- het_mean - 3 * het_sd   # feel free to adjust
het_threshold_high <- het_mean + 3 * het_sd   # feel free to adjust
miss_threshold     <- ___FILL_IN___             # pick a value by eye from the plot above

qc <- qc %>%
  mutate(fail_het  = het_rate < het_threshold_low | het_rate > het_threshold_high,
         fail_miss = F_MISS > miss_threshold,
         fail_qc   = fail_het | fail_miss)

table(fail_het = qc$fail_het, fail_miss = qc$fail_miss)
```

Re-draw the plot, colouring the samples you're about to remove, to sanity
check your choice before acting on it:

```{r}
ggplot(qc, aes(x = F_MISS, y = het_rate, color = fail_qc)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_x_log10() +
  scale_color_manual(values = c(`FALSE` = "grey40", `TRUE` = "firebrick")) +
  labs(color = "Fails QC", title = "Samples flagged for removal") +
  theme_minimal(base_size = 12)
```

Happy with it? Write the fail list out in the two-column format PLINK
expects for `--remove` (FID, IID, no header):

```{r}
qc %>%
  filter(fail_qc) %>%
  select(FID, IID) %>%
  write.table("results/qc_fail_samples.txt",
              row.names = FALSE, col.names = FALSE, quote = FALSE, sep = "\t")

cat("Flagged", sum(qc$fail_qc), "of", nrow(qc), "samples for removal.\n")
```

## Step 5 — apply the removal (and a standard SNP-missingness filter)

While we're here, also drop SNPs with a failing call rate genome-wide
(`--geno 0.05` = drop any SNP missing in more than 5% of remaining
samples) — this is normally a quick, non-controversial default rather than
something you eyeball a plot for, so we apply it directly:

```{bash}
plink2 --bfile data/study \
  --remove results/qc_fail_samples.txt \
  --geno 0.05 \
  --make-bed \
  --out data/study_qc1
```

```{bash}
wc -l data/study_qc1.fam
wc -l data/study_qc1.bim
```

You should have fewer samples than you started with, and fewer SNPs than
50,000 (some fraction of the SNPs on "chromosome 9" were deliberately
built to fail this exact filter — you'll likely notice that when you look
at what got dropped).

**Ancestry is still unresolved at this point** — nothing here removed
samples for being a different ancestry; only for missingness/heterozygosity.
That's what notebook 02 is for.
