---
title: "02 — PCA & Ancestry"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# The setup chunk runs in this file's own directory; every *later*
# chunk runs from the project root thanks to root.dir below.
proj_root <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj_root)
library(dplyr)
library(ggplot2)
source(file.path(proj_root, "R/sim_utils.R"))
source(file.path(proj_root, "R/plot_utils.R"))
```

## Why compare to a reference panel

Clustering your own samples in isolation can tell you "these people don't
look like the rest," but not *who they actually are* — and in a real study
you often want to know which continental/regional ancestry group a cluster
corresponds to, or confirm that your "cases" and "controls" weren't
recruited from subtly different populations. The standard approach:
combine your samples with a panel of individuals from known reference
populations (in real studies, often 1000 Genomes or HapMap), run PCA
together, and see which reference population each of your clusters lines
up with.

`data/reference` is exactly that kind of panel here: several groups of
individuals from **known** populations (labels included), generated
independently of your study samples.

## Step 1 — merge your (sample-QC'd) data with the reference panel

Your study data and the reference panel were simulated from the *identical*
SNP map, so merging is just a matter of stacking individuals — no strand
flips or allele mismatches to worry about (real merges usually aren't this
clean; `merge_same_snps()` below is a shortcut that only works because we
guaranteed an identical marker set).

```{r}
merge_same_snps("data/study_qc1", "data/reference", "results/study_plus_ref")
```

```{bash}
wc -l results/study_plus_ref.fam
```

## Step 2 — run PCA on the combined set

```{bash}
plink2 --bfile results/study_plus_ref \
  --maf 0.05 \
  --indep-pairwise 200 50 0.2 \
  --out results/pca_prune
plink2 --bfile results/study_plus_ref \
  --extract results/pca_prune.prune.in \
  --pca 10 \
  --out results/pca
```

Two things worth noticing in that command: we filtered to reasonably
common SNPs (`--maf 0.05`) and LD-pruned (`--indep-pairwise`) before PCA.
Rare variants and SNPs in tight linkage disequilibrium can distort the
top principal components — pruning keeps the PCs reflecting broad
ancestry structure rather than a handful of correlated local SNPs.

Check how many SNPs pruning actually removed here:

```{bash}
wc -l < results/pca_prune.prune.in
wc -l < results/pca_prune.prune.out
```

You'll find it removed essentially **nothing**. That's a property of this
simulation, not of the method: every SNP here was generated independently,
so there is no linkage disequilibrium for pruning to find. On real data
this step routinely discards half your markers or more, and skipping it
genuinely does distort the top PCs. Keep the step in your muscle memory —
notebook 05 comes back to what LD's absence changes about this dataset.

## Step 3 — plot PC1 vs PC2, reference populations labeled

```{r}
eigenvec <- data.table::fread("results/pca.eigenvec")
names(eigenvec) <- sub("^#", "", names(eigenvec))

# Reference samples' IDs start with "R"; everyone else is a study sample.
eigenvec <- eigenvec %>%
  mutate(group = ifelse(grepl("^R", IID), "Reference (labeled)", "Study sample (unlabeled)"))

# Note: the reference .fam alone doesn't carry population names. For this
# exercise we only need to see that reference individuals form distinct,
# labeled clusters; a real workflow would carry population labels through
# a companion metadata file (e.g. 1000 Genomes' panel file).

ggplot(eigenvec, aes(x = PC1, y = PC2, color = group)) +
  geom_point(alpha = 0.6, size = 1) +
  labs(title = "PCA: study samples vs. reference panel") +
  theme_minimal(base_size = 12)
```

**🔧 YOUR TURN:** rerun the plot above but facet or color by something
more informative than just "reference vs. study" — e.g. plot PC1 vs PC2
with reference points colored by their FID prefix pattern, so you can see
whether the reference panel actually forms multiple distinct clusters (it
should — there are 4 reference populations in there, and one of them
doesn't match anything in your study data at all):

```{r}
# Reference FIDs look like "R00001".."R00320", in 4 contiguous blocks of 80.
eigenvec <- eigenvec %>%
  mutate(ref_pop_guess = case_when(
    !grepl("^R", IID) ~ NA_character_,
    TRUE ~ ___FILL_IN___   # derive a 4-level label from IID's row order or numeric suffix
  ))

ggplot(eigenvec, aes(x = PC1, y = PC2)) +
  geom_point(data = filter(eigenvec, is.na(ref_pop_guess)), color = "grey80", size = 1, alpha = 0.5) +
  geom_point(data = filter(eigenvec, !is.na(ref_pop_guess)), aes(color = ref_pop_guess), size = 1.4) +
  labs(title = "Study samples (grey) against 4 labeled reference populations") +
  theme_minimal(base_size = 12)
```

## Step 4 — identify and remove ancestry outliers

Now the real question: which of *your study samples* sit with the
majority reference cluster, and which sit off to the side with a minority
reference population? Decide on a rule (distance from the main cluster's
centroid, a manual boundary drawn from the plot, k-means on PC1/PC2 — pick
whichever you find easiest to justify) and flag the outliers.

```{r}
study_pts <- eigenvec %>% filter(is.na(ref_pop_guess))

# A simple, explainable approach: define the majority cluster's centroid
# from whichever group of study points is largest/densest, then flag
# anyone beyond some distance from it.
centroid_pc1 <- ___FILL_IN___   # e.g. median(PC1) of the obvious majority cluster
centroid_pc2 <- ___FILL_IN___
dist_threshold <- ___FILL_IN___  # pick from the plot

study_pts <- study_pts %>%
  mutate(dist_from_centroid = sqrt((PC1 - centroid_pc1)^2 + (PC2 - centroid_pc2)^2),
         ancestry_outlier = dist_from_centroid > dist_threshold)

table(study_pts$ancestry_outlier)
```

```{r}
ggplot(study_pts, aes(PC1, PC2, color = ancestry_outlier)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c(`FALSE` = "grey40", `TRUE` = "firebrick")) +
  labs(title = "Study samples flagged as ancestry outliers") +
  theme_minimal(base_size = 12)
```

Write the outlier list and apply it:

```{r}
study_pts %>%
  filter(ancestry_outlier) %>%
  select(FID, IID) %>%
  write.table("results/ancestry_outliers.txt",
              row.names = FALSE, col.names = FALSE, quote = FALSE, sep = "\t")
```

```{bash}
plink2 --bfile data/study_qc1 \
  --remove results/ancestry_outliers.txt \
  --make-bed \
  --out data/study_qc2
wc -l data/study_qc2.fam
```

If your threshold was reasonable, `data/study_qc2` should now be close to
a single, ancestrally homogeneous group — that's the file the rest of the
workshop builds on for the "corrected" analysis. Keep `data/study_qc1`
around too; notebook 04 deliberately reuses it to show what happens when
you *skip* this step.
