---
title: "02 — PCA & Ancestry"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Locate the project root and work from there. This makes the notebook behave
# the same whether you press Knit or run chunks one at a time, and whether
# RStudio's working directory is the project root or this notebooks/ folder.
proj_root <- if (file.exists("R/sim_utils.R")) normalizePath(".") else normalizePath("..")
if (!file.exists(file.path(proj_root, "R", "sim_utils.R"))) {
  stop("Can't find the project root from ", getwd(), ".\n",
       "Fix: in the Console, run  setwd(\"/workspaces/popgen-gwas-tutorial\")  ",
       "then run this chunk again. (Or: click popgen-gwas-tutorial.Rproj in the ",
       "Files pane to reopen as a project, which sets this automatically from now on.)")
}
knitr::opts_knit$set(root.dir = proj_root)                     # applies when knitting
if (!isTRUE(getOption("knitr.in.progress"))) setwd(proj_root)  # applies when running chunks
library(dplyr)
library(ggplot2)
source(file.path(proj_root, "R/sim_utils.R"))
source(file.path(proj_root, "R/plot_utils.R"))
```

## Why compare to a reference panel

Clustering your own samples in isolation can tell you "these people don't
look like the rest," but not *who they actually are* — and in a real study
you often want to know which continental/regional ancestry group a cluster
corresponds to, or confirm that your "cases" and "controls" weren't
recruited from subtly different populations. The standard approach:
combine your samples with a panel of individuals from known reference
populations (in real studies, often 1000 Genomes or HapMap), run PCA
together, and see which reference population each of your clusters lines
up with.

`data/reference` is exactly that kind of panel here: several groups of
individuals from **known** populations (labels included), generated
independently of your study samples.

## Step 1 — merge your (sample-QC'd) data with the reference panel

Your study data and the reference panel were simulated from the *identical*
SNP map, so merging is just a matter of stacking individuals — no strand
flips or allele mismatches to worry about (real merges usually aren't this
clean; `merge_same_snps()` below is a shortcut that only works because we
guaranteed an identical marker set).

```{bash}
mkdir -p results
```

```{r}
merge_same_snps("data/study_qc1", "data/reference", "results/study_plus_ref")
```

```{bash}
wc -l results/study_plus_ref.fam
```

## Step 2 — run PCA on the combined set

```{bash}
plink2 --bfile results/study_plus_ref \
  --maf 0.05 \
  --indep-pairwise 200 50 0.2 \
  --out results/pca_prune
plink2 --bfile results/study_plus_ref \
  --extract results/pca_prune.prune.in \
  --pca 10 \
  --out results/pca
```

Two things worth noticing in that command: we filtered to reasonably
common SNPs (`--maf 0.05`) and LD-pruned (`--indep-pairwise`) before PCA.
Rare variants and SNPs in tight linkage disequilibrium can distort the
top principal components — pruning keeps the PCs reflecting broad
ancestry structure rather than a handful of correlated local SNPs.

Check how many SNPs pruning actually removed here:

```{bash}
wc -l < results/pca_prune.prune.in
wc -l < results/pca_prune.prune.out
```

You'll find it removed only a small handful — nothing like the "routinely
discards half your markers or more" you'd see on real data. That's a
property of this simulation: almost every SNP here was generated
independently, with no correlation to its neighbors at all. The few that
*do* get pruned aren't scattered randomly, though — they cluster tightly
around a small number of specific positions. Notebook 05 comes back to
exactly where, and why those particular spots are the exception.

## Step 3 — plot PC1 vs PC2, reference populations labeled

```{r}
eigenvec <- data.table::fread("results/pca.eigenvec")
names(eigenvec) <- sub("^#", "", names(eigenvec))

# Reference samples' IDs start with "R"; everyone else is a study sample.
eigenvec <- eigenvec %>%
  mutate(group = ifelse(grepl("^R", IID), "Reference (labeled)", "Study sample (unlabeled)"))

# Note: the reference .fam alone doesn't carry population names. For this
# exercise we only need to see that reference individuals form distinct,
# labeled clusters; a real workflow would carry population labels through
# a companion metadata file (e.g. 1000 Genomes' panel file).

ggplot(eigenvec, aes(x = PC1, y = PC2, color = group)) +
  geom_point(alpha = 0.6, size = 1) +
  labs(title = "PCA: study samples vs. reference panel") +
  theme_minimal(base_size = 12)
```

**🔧 YOUR TURN:** rerun the plot above but facet or color by something
more informative than just "reference vs. study" — e.g. plot PC1 vs PC2
with reference points colored by their FID prefix pattern, so you can see
whether the reference panel actually forms multiple distinct clusters (it
should — there are 4 reference populations in there, and one of them
doesn't match anything in your study data at all):

```{r}
# Reference FIDs look like "R00001".."R00320", in 4 contiguous blocks of 80.
eigenvec <- eigenvec %>%
  mutate(ref_pop_guess = case_when(
    !grepl("^R", IID) ~ NA_character_,
    TRUE ~ ___FILL_IN___   # derive a 4-level label from IID's row order or numeric suffix
  ))

ggplot(eigenvec, aes(x = PC1, y = PC2)) +
  geom_point(data = filter(eigenvec, is.na(ref_pop_guess)), color = "grey80", size = 1, alpha = 0.5) +
  geom_point(data = filter(eigenvec, !is.na(ref_pop_guess)), aes(color = ref_pop_guess), size = 1.4) +
  labs(title = "Study samples (grey) against 4 labeled reference populations") +
  theme_minimal(base_size = 12)
```

## Step 4 — identify and remove ancestry outliers

Now the real question: which of *your study samples* sit with the
majority reference cluster, and which sit off to the side with a minority
reference population? Decide on a rule (distance from the main cluster's
centroid, a manual boundary drawn from the plot, k-means on PC1/PC2 — pick
whichever you find easiest to justify) and flag the outliers.

```{r}
study_pts <- eigenvec %>% filter(is.na(ref_pop_guess))

# A simple, explainable approach: define the majority cluster's centroid
# from whichever group of study points is largest/densest, then flag
# anyone beyond some distance from it.
centroid_pc1 <- ___FILL_IN___   # e.g. median(PC1) of the obvious majority cluster
centroid_pc2 <- ___FILL_IN___
dist_threshold <- ___FILL_IN___  # pick from the plot

study_pts <- study_pts %>%
  mutate(dist_from_centroid = sqrt((PC1 - centroid_pc1)^2 + (PC2 - centroid_pc2)^2),
         ancestry_outlier = dist_from_centroid > dist_threshold)

table(study_pts$ancestry_outlier)
```

```{r}
ggplot(study_pts, aes(PC1, PC2, color = ancestry_outlier)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c(`FALSE` = "grey40", `TRUE` = "firebrick")) +
  labs(title = "Study samples flagged as ancestry outliers") +
  theme_minimal(base_size = 12)
```

Write the outlier list and apply it:

```{r}
study_pts %>%
  filter(ancestry_outlier) %>%
  select(FID, IID) %>%
  write.table("results/ancestry_outliers.txt",
              row.names = FALSE, col.names = FALSE, quote = FALSE, sep = "\t")
```

```{bash}
plink2 --bfile data/study_qc1 \
  --remove results/ancestry_outliers.txt \
  --make-bed \
  --out data/study_qc2
wc -l data/study_qc2.fam
```

If your threshold was reasonable, `data/study_qc2` should now be close to
a single, ancestrally homogeneous group — that's the file the rest of the
workshop builds on for the "corrected" analysis. Keep `data/study_qc1`
around too; notebook 04 deliberately reuses it to show what happens when
you *skip* this step.
