---
title: "03 — MAF & HWE Filtering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Locate the project root and work from there. This makes the notebook behave
# the same whether you press Knit or run chunks one at a time, and whether
# RStudio's working directory is the project root or this notebooks/ folder.
proj_root <- if (file.exists("R/sim_utils.R")) normalizePath(".") else normalizePath("..")
if (!file.exists(file.path(proj_root, "R", "sim_utils.R"))) {
  stop("Can't find the project root from ", getwd(), ".\n",
       "Fix: in the Console, run  setwd(\"/workspaces/popgen-gwas-tutorial\")  ",
       "then run this chunk again. (Or: click popgen-gwas-tutorial.Rproj in the ",
       "Files pane to reopen as a project, which sets this automatically from now on.)")
}
knitr::opts_knit$set(root.dir = proj_root)                     # applies when knitting
if (!isTRUE(getOption("knitr.in.progress"))) setwd(proj_root)  # applies when running chunks
library(dplyr)
library(ggplot2)
```

With sample QC and ancestry outliers dealt with (`data/study_qc2`), the
last two variant-level filters before running the association test:
**minor allele frequency (MAF)** and **Hardy-Weinberg equilibrium (HWE)**.

## Why filter on MAF

Very rare variants give you very little statistical power (few or no
copies of the minor allele among your cases and controls to compare), and
at the extreme, genotyping errors are more likely to masquerade as a real
rare allele than a common one. A standard case/control GWAS filter is
somewhere around 1% MAF — low enough to keep genuinely interesting
variants, high enough to drop noise.

```{bash}
mkdir -p results
plink2 --bfile data/study_qc2 --freq --out results/study_qc2_freq
```

```{r}
freq <- data.table::fread("results/study_qc2_freq.afreq")
names(freq) <- sub("^#", "", names(freq))
freq <- freq %>% mutate(maf = pmin(ALT_FREQS, 1 - ALT_FREQS))

ggplot(freq, aes(x = maf)) +
  geom_histogram(bins = 60, fill = "#4C72B0") +
  labs(title = "Minor allele frequency, post ancestry-QC", x = "MAF") +
  theme_minimal(base_size = 12)
```

**🔧 YOUR TURN:** pick a MAF threshold from that distribution and check how
many SNPs it would remove before committing to it:

```{r}
maf_threshold <- ___FILL_IN___   # e.g. try 0.01 and see how many SNPs that drops

sum(freq$maf < maf_threshold)
mean(freq$maf < maf_threshold)
```

## Why test HWE — and who exactly gets tested

Under random mating, genotype counts at a neutral SNP should follow
Hardy-Weinberg proportions ($p^2, 2pq, q^2$). A SNP that deviates sharply
is often a technical artefact (a genotyping/calling error) rather than
biology — which is exactly what we injected on chromosome 4.

There's a subtlety here that matters, and it's a good example of why you
read your tool's documentation instead of assuming. At a SNP that *truly*
affects disease risk, cases are **expected** to deviate from HWE — that's
not an error, it's a direct consequence of the SNP influencing who becomes
a case. So the textbook advice is to test HWE **in controls only**, to
avoid penalizing your best real hits.

PLINK **1.9** did exactly that automatically for case/control data. **PLINK
2 does not** — it computes HWE over all founders (cases and controls
together), as its own log output will tell you if you read it. Run it and
look at what PLINK actually reports:

```{bash}
plink2 --bfile data/study_qc2 --hardy --out results/study_qc2_hwe
grep -iE "founders|hardy" results/study_qc2_hwe.log
```

Note the "(founders only)" in that message — **not** "controls only". If
you want the controls-only test in PLINK 2, you have to ask for it
explicitly, by restricting to controls first:

```{bash}
awk '$6==1 {print $1"\t"$2}' data/study_qc2.fam > results/controls_only.txt
plink2 --bfile data/study_qc2 --keep results/controls_only.txt \
  --hardy --out results/study_qc2_hwe_controls
```

**🔧 YOUR TURN:** do the two versions actually disagree on this dataset?

```{r}
hwe_all  <- data.table::fread("results/study_qc2_hwe.hardy")
hwe_ctrl <- data.table::fread("results/study_qc2_hwe_controls.hardy")
names(hwe_all)  <- sub("^#", "", names(hwe_all))
names(hwe_ctrl) <- sub("^#", "", names(hwe_ctrl))

cmp <- merge(hwe_all[, .(ID, P_founders = P)],
             hwe_ctrl[, .(ID, P_controls = P)], by = "ID")
# how many SNPs would each approach flag at p < 1e-6?
c(founders = sum(cmp$P_founders < 1e-6, na.rm = TRUE),
  controls = sum(cmp$P_controls < 1e-6, na.rm = TRUE))
```

On *this* dataset the two give nearly the same answer, because the causal
SNPs' effects are modest enough that case-driven HWE deviation stays small
— you can check that none of your eventual hits get filtered out either
way. In a study with larger effect sizes, the difference is exactly where
you'd risk throwing away a real finding. We'll use PLINK 2's default
below; just know what it's doing.

```{r}
hwe <- data.table::fread("results/study_qc2_hwe.hardy")
names(hwe) <- sub("^#", "", names(hwe))
head(hwe)
```

```{r}
ggplot(hwe, aes(x = -log10(P))) +
  geom_histogram(bins = 60, fill = "#8C8C8C") +
  labs(title = "HWE test p-values across all SNPs") +
  theme_minimal(base_size = 12)
```

**🔧 YOUR TURN:** most SNPs should show no HWE deviation at all (p-values
close to uniform). A cluster of very small p-values is worth investigating
before you blindly filter — look at *which chromosome* the most extreme
hits fall on:

```{r}
hwe %>%
  filter(P < 1e-6) %>%
  count(CHROM) %>%
  arrange(desc(n))
```

Does that match what notebook 00 told you about where a deliberate
genotyping artefact was built in? Pick an HWE p-value threshold:

```{r}
hwe_threshold <- ___FILL_IN___   # a common default is 1e-6
```

## Apply both filters together

```{bash}
plink2 --bfile data/study_qc2 \
  --maf 0.01 \
  --hwe 1e-6 \
  --make-bed \
  --out data/study_final
wc -l data/study_final.bim
```

*(Replace the `--maf`/`--hwe` values above with the thresholds you settled
on, if different.)*

`data/study_final` is the fully QC'd dataset the rest of the workshop
calls the "corrected" analysis.

## A preview of something you'll see in notebook 04

Out of curiosity, try running the exact same `--hardy` command on
`data/study_qc1` (sample-QC'd, but *before* ancestry outliers were
removed) instead of `data/study_qc2`, and look specifically at chromosome
3:

```{bash}
plink2 --bfile data/study_qc1 --hardy --out results/study_qc1_hwe
```

```{r}
hwe_naive <- data.table::fread("results/study_qc1_hwe.hardy")
names(hwe_naive) <- sub("^#", "", names(hwe_naive))
hwe_naive %>% filter(CHROM == 3) %>% arrange(P) %>% head(3)
```

You should find one SNP on chromosome 3 with an almost impossibly small
HWE p-value in the ancestry-mixed data — far more extreme than the
genuine artefact block on chromosome 4. This is the **Wahlund effect**:
pooling two genetically differentiated subpopulations that each have very
different allele frequencies *at one particular SNP* creates an apparent
massive excess of homozygotes there, purely from mixing structured
groups — no genotyping error required. It's a real and important failure
mode: a naive HWE filter applied to ancestry-mixed data wouldn't just
give you a spurious warning here, it would actually **delete this SNP
before you ever got to test it for association** — which is exactly why
the "naive" analysis in notebook 04 deliberately does *not* apply an HWE
filter, so you can see what that SNP does when it survives to the
association test uncorrected. Keep this chromosome number in mind.
