---
title: "03 — MAF & HWE Filtering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
library(dplyr)
library(ggplot2)
```

With sample QC and ancestry outliers dealt with (`data/study_qc2`), the
last two variant-level filters before running the association test:
**minor allele frequency (MAF)** and **Hardy-Weinberg equilibrium (HWE)**.

## Why filter on MAF

Very rare variants give you very little statistical power (few or no
copies of the minor allele among your cases and controls to compare), and
at the extreme, genotyping errors are more likely to masquerade as a real
rare allele than a common one. A standard case/control GWAS filter is
somewhere around 1% MAF — low enough to keep genuinely interesting
variants, high enough to drop noise.

```{bash}
mkdir -p results
plink2 --bfile data/study_qc2 --freq --out results/study_qc2_freq
```

```{r}
freq <- data.table::fread("results/study_qc2_freq.afreq")
names(freq) <- sub("^#", "", names(freq))
freq <- freq %>% mutate(maf = pmin(ALT_FREQS, 1 - ALT_FREQS))

ggplot(freq, aes(x = maf)) +
  geom_histogram(bins = 60, fill = "#4C72B0") +
  labs(title = "Minor allele frequency, post ancestry-QC", x = "MAF") +
  theme_minimal(base_size = 12)
```

**🔧 YOUR TURN:** pick a MAF threshold from that distribution and check how
many SNPs it would remove before committing to it:

```{r}
maf_threshold <- ___FILL_IN___   # e.g. try 0.01 and see how many SNPs that drops

sum(freq$maf < maf_threshold)
mean(freq$maf < maf_threshold)
```

## Why test HWE — and why only in controls

Under random mating, genotype counts at a neutral SNP should follow
Hardy-Weinberg proportions ($p^2, 2pq, q^2$). A SNP that deviates sharply
is often a technical artefact (a genotyping/calling error) rather than
biology — which is exactly what we injected on chromosome 4.

**Important:** HWE is tested in **controls only** by default whenever
PLINK sees a case/control phenotype loaded (cases are allowed to deviate
from HWE *at a truly associated SNP* — that's not an error, it's a
consequence of the SNP affecting who becomes a case. Testing HWE across
everyone would penalize your best hits). Confirm PLINK is doing this:

```{bash}
plink2 --bfile data/study_qc2 --hardy --out results/study_qc2_hwe
```

```{r}
hwe <- data.table::fread("results/study_qc2_hwe.hardy")
names(hwe) <- sub("^#", "", names(hwe))
head(hwe)
```

```{r}
ggplot(hwe, aes(x = -log10(P))) +
  geom_histogram(bins = 60, fill = "#8C8C8C") +
  labs(title = "HWE test p-values across all SNPs") +
  theme_minimal(base_size = 12)
```

**🔧 YOUR TURN:** most SNPs should show no HWE deviation at all (p-values
close to uniform). A cluster of very small p-values is worth investigating
before you blindly filter — look at *which chromosome* the most extreme
hits fall on:

```{r}
hwe %>%
  filter(P < 1e-6) %>%
  count(CHROM) %>%
  arrange(desc(n))
```

Does that match what notebook 00 told you about where a deliberate
genotyping artefact was built in? Pick an HWE p-value threshold:

```{r}
hwe_threshold <- ___FILL_IN___   # a common default is 1e-6
```

## Apply both filters together

```{bash}
plink2 --bfile data/study_qc2 \
  --maf 0.01 \
  --hwe 1e-6 \
  --make-bed \
  --out data/study_final
wc -l data/study_final.bim
```

*(Replace the `--maf`/`--hwe` values above with the thresholds you settled
on, if different.)*

`data/study_final` is the fully QC'd dataset the rest of the workshop
calls the "corrected" analysis.

## A preview of something you'll see in notebook 04

Out of curiosity, try running the exact same `--hardy` command on
`data/study_qc1` (sample-QC'd, but *before* ancestry outliers were
removed) instead of `data/study_qc2`, and look specifically at chromosome
3:

```{bash}
plink2 --bfile data/study_qc1 --hardy --out results/study_qc1_hwe
```

```{r}
hwe_naive <- data.table::fread("results/study_qc1_hwe.hardy")
names(hwe_naive) <- sub("^#", "", names(hwe_naive))
hwe_naive %>% filter(CHROM == 3) %>% arrange(P) %>% head(3)
```

You should find one SNP on chromosome 3 with an almost impossibly small
HWE p-value in the ancestry-mixed data — far more extreme than the
genuine artefact block on chromosome 4. This is the **Wahlund effect**:
pooling two genetically differentiated subpopulations that each have very
different allele frequencies *at one particular SNP* creates an apparent
massive excess of homozygotes there, purely from mixing structured
groups — no genotyping error required. It's a real and important failure
mode: a naive HWE filter applied to ancestry-mixed data wouldn't just
give you a spurious warning here, it would actually **delete this SNP
before you ever got to test it for association** — which is exactly why
the "naive" analysis in notebook 04 deliberately does *not* apply an HWE
filter, so you can see what that SNP does when it survives to the
association test uncorrected. Keep this chromosome number in mind.
