---
title: "04 — Running the GWAS: Naive vs. Ancestry-Corrected"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Locate the project root and work from there. This makes the notebook behave
# the same whether you press Knit or run chunks one at a time, and whether
# RStudio's working directory is the project root or this notebooks/ folder.
proj_root <- if (file.exists("R/sim_utils.R")) normalizePath(".") else normalizePath("..")
if (!file.exists(file.path(proj_root, "R", "sim_utils.R"))) {
  stop("Can't find the project root from ", getwd(), ".\n",
       "Fix: in the Console, run  setwd(\"/workspaces/popgen-gwas-tutorial\")  ",
       "then run this chunk again. (Or: click popgen-gwas-tutorial.Rproj in the ",
       "Files pane to reopen as a project, which sets this automatically from now on.)")
}
knitr::opts_knit$set(root.dir = proj_root)                     # applies when knitting
if (!isTRUE(getOption("knitr.in.progress"))) setwd(proj_root)  # applies when running chunks
library(dplyr)
library(ggplot2)
source(file.path(proj_root, "R/plot_utils.R"))
```

This is the centerpiece of the workshop: we're going to run the **same**
association test twice, on two versions of the data that differ in
exactly one way — whether ancestry outliers were removed — and see what
population stratification actually does to a GWAS.

- **Naive**: `data/study_qc1` (sample QC done, ancestry outliers *not*
  removed) — filtered on MAF/HWE fresh, right now, in this notebook.
- **Corrected**: `data/study_final` (the fully QC'd file from notebook 03).

## Branch 1 — the naive analysis

Deliberately **not** applying `--hwe` here — recall from notebook 03 that
one particular SNP fails HWE catastrophically the moment ancestries are
pooled (the Wahlund effect). Filtering on HWE first would delete it
before we ever got to see what it does in an association test. A naive
analyst who ran HWE QC on correctly-separated ancestries wouldn't have
this problem — but one who pooled ancestries first, the way this "naive"
branch does, might not even get that warning sign before publishing a
false result. MAF filtering alone doesn't have this issue, so we keep it.

```{bash}
mkdir -p results
plink2 --bfile data/study_qc1 --maf 0.01 --make-bed --out data/study_qc1_filtered
```

```{bash}
plink2 --bfile data/study_qc1_filtered \
  --glm allow-no-covars \
  --out results/gwas_naive
```

## Branch 2 — the ancestry-corrected analysis

```{bash}
plink2 --bfile data/study_final \
  --glm allow-no-covars \
  --out results/gwas_corrected
```

`--glm` writes a `.PHENO1.glm.logistic.hybrid` file (or `.glm.logistic`,
depending on your PLINK2 build — check the `results/` folder if the exact
name below doesn't match and adjust the path).

```{r}
naive     <- read_glm(Sys.glob("results/gwas_naive.*.glm.logistic*")[1])
corrected <- read_glm(Sys.glob("results/gwas_corrected.*.glm.logistic*")[1])
nrow(naive); nrow(corrected)
```

## Compare: Manhattan plots side by side

A reasonable multiple-testing threshold for ~25,000 independent tests is
roughly $0.05/25000 \approx 2\times10^{-6}$ — looser than the $5\times10^{-8}$
you'll see quoted for real ~1-million-SNP genotyping arrays, because we
ran far fewer tests here. The threshold should always reflect how many
tests you actually did.

```{r, fig.height=4}
sig_line <- 2e-6
plot_manhattan(naive, title = "Naive (ancestry outliers included)", sig_line = sig_line)
plot_manhattan(corrected, title = "Ancestry-corrected", sig_line = sig_line)
```

**🔧 YOUR TURN:** find the SNP(s) that cross the significance line in one
analysis but not the other — check *both* directions, because ancestry
correction can both remove a false signal and reveal a true one that was
previously drowned out:

```{r}
sig_naive     <- naive     %>% filter(P < sig_line) %>% select(ID, CHROM, POS, P, OR)
sig_corrected <- corrected %>% filter(P < sig_line) %>% select(ID, CHROM, POS, P, OR)

only_in_naive     <- anti_join(sig_naive, sig_corrected, by = "ID")
only_in_corrected <- anti_join(sig_corrected, sig_naive, by = "ID")
only_in_naive
only_in_corrected
```

The SNP that's significant only in the naive analysis is not a mistake in
the pipeline — it's doing exactly what a real confounded,
population-stratified false positive does (and it's on the chromosome
notebook 03 flagged with the extreme HWE artefact — that's not a
coincidence).

The SNP that's significant only *after* correction is a different kind of
lesson, and a subtler one: it sits immediately next to a real causal SNP,
tagging along on that SNP's true signal via linkage disequilibrium
(notebook 05 digs into this properly). In the naive analysis it's just
short of the line; removing the stratification noise nudges it over.
Nothing was manufactured here — the underlying signal was real in both
analyses, just partly buried by extra noise in one of them. That's the
general shape of why correcting a real confound can occasionally *reveal*
a hit as well as remove one: less noise means more of what's genuinely
there becomes visible, in both directions at once.

## Compare: QQ plots and genomic inflation (λ)

```{r, fig.height=4}
plot_qq(naive$P, title = paste0("Naive — λ_GC = ", round(lambda_gc(naive$P), 3)))
plot_qq(corrected$P, title = paste0("Corrected — λ_GC = ", round(lambda_gc(corrected$P), 3)))
```

$\lambda_{GC}$ close to 1 means the bulk of your test statistics (almost
all of which should be null — most SNPs have no real effect) behave the
way chance alone predicts. A value well above 1, with the QQ plot bowing
up away from the diagonal early, is the classic quantitative signature of
unaccounted-for population stratification — the same problem you can see
visually in the Manhattan plot, but as a single number you could report in
a paper.

**🔧 YOUR TURN:** report both λ_GC values here and describe, in a
sentence or two, what changed between the two analyses and why:

```{r}
cat("Naive lambda_GC:    ", round(lambda_gc(naive$P), 3), "\n")
cat("Corrected lambda_GC:", round(lambda_gc(corrected$P), 3), "\n")
```

_Your answer:_ ___FILL_IN___

## What's next

`05_interpretation_manhattan_qq.Rmd` digs into the **corrected** results
specifically: which hits reached significance, how their estimated effect
sizes compare to expectations, and why a couple of real signals might
still be missing.
