---
title: "05 — Interpreting the GWAS Hits"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# The setup chunk runs in this file's own directory; every *later*
# chunk runs from the project root thanks to root.dir below.
proj_root <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj_root)
library(dplyr)
library(ggplot2)
source(file.path(proj_root, "R/plot_utils.R"))
```

We'll work only with the **ancestry-corrected** results from notebook 04
(`results/gwas_corrected`) from here on — this is the analysis you'd
actually stand behind.

```{r}
corrected <- read_glm(Sys.glob("results/gwas_corrected.*.glm.logistic*")[1])
sig_line <- 2e-6
```

## Which SNPs reached significance?

```{r}
hits <- corrected %>%
  filter(P < sig_line) %>%
  arrange(P) %>%
  select(ID, CHROM, POS, A1, OR, P)
hits
```

For each hit, sanity-check the reported odds ratio: **which allele is it
relative to?** The `A1` column tells you which allele PLINK tested. If an
`OR` comes out less than 1, that means the *other* allele (not `A1`) is
the one associated with higher risk — take the reciprocal (`1/OR`) to
express the effect in terms of the risk-increasing allele. (On this
dataset's seed, every current hit happens to already have `OR > 1`, so the
line below is a no-op here — but don't skip it: which allele lands as A1
is an upstream bookkeeping choice, not something you control, and it
won't always land this conveniently. This is a real and common gotcha when
reading any GWAS summary statistics table.)

```{r}
hits <- hits %>%
  mutate(OR_risk_allele = ifelse(OR < 1, 1 / OR, OR))
hits
```

## How many chromosomes show a real peak?

```{r, fig.height=4}
plot_manhattan(corrected, title = "Corrected GWAS", sig_line = sig_line)
```

**🔧 YOUR TURN:** count how many *distinct* SNPs, and how many *distinct*
chromosomes, have at least one hit crossing the significance line:

```{r}
hits %>% distinct(CHROM)
nrow(hits)
```

Recall from notebook 00 that this dataset embeds SNPs across a **range**
of allele frequencies and effect sizes, deliberately chosen so that not
every real signal would be expected to reach significance at this sample
size. You should find **4 significant SNPs across 2 chromosomes** — notice
those two numbers aren't the same. More than one real hit can share a
chromosome (chromosome 2 has two here), which is completely realistic:
different genes at different positions on the same chromosome can each be
independently associated with a disease. Counting distinct *SNPs* is what
actually matters; counting distinct chromosomes just happens to
undercount when that overlap occurs. And there are more than 4 real
causal variants in this data in the first place — the ones you don't see
at all are not a failure of your analysis, they're underpowered. That gap
between "how many real effects exist" and "how many you can detect" is
the single most important practical fact about GWAS, and this notebook's
discussion questions come back to it.

## A closer look: zoom into the top chromosome

```{r}
top_chr <- hits$CHROM[1]

corrected %>%
  filter(CHROM == top_chr) %>%
  ggplot(aes(x = POS, y = -log10(P))) +
  geom_point(alpha = 0.6, size = 1) +
  geom_hline(yintercept = -log10(sig_line), linetype = "dashed", color = "firebrick") +
  labs(title = paste("Chromosome", top_chr, "region"), x = "Position") +
  theme_minimal(base_size = 12)
```

Notice **two separate peaks** on this one chromosome, not one — two
different causal SNPs happen to have been placed here. Look closely at
either peak individually, though, and each one is a **single isolated
point**: its immediate neighbours sit right down in the noise, with
nothing gradually building up or tailing off around it. That's a
deliberate property of this simulation, and it's worth pausing on, because
it is *not* what real GWAS data looks like.

Every SNP in this dataset was simulated independently, so there is no
**linkage disequilibrium (LD)** — no correlation between neighbouring
markers. In real data, a causal variant drags its neighbours up with it:
SNPs physically close by are inherited together, so they carry a partial
copy of the same signal and form the characteristic "skyscraper with
shoulders" shape you see in published Manhattan plots. That's also why
real GWAS hits are reported as *loci* rather than single variants, and why
the top SNP at a locus is very often not the causal one — it's just the
best-tagged marker nearby. Multiple real, independent loci sitting on the
same chromosome (like the two peaks above) is a separate and unrelated
phenomenon from LD — LD is about correlation *within* one locus's
neighbourhood, not about how many independent loci a chromosome can hold.

Two things in this workshop are quietly shaped by the absence of LD: the
LD-pruning step in notebook 02 had almost nothing to remove, and our
significance threshold treats all ~25,000 SNPs as independent tests, which
here is actually true (in real data it's conservative, because correlated
SNPs aren't independent tests).

Compare these isolated peaks to what a chromosome with *no* real signal
looks like:

```{r}
corrected %>%
  filter(CHROM == 5) %>%   # chromosome 5 was built with no injected signal at all
  ggplot(aes(x = POS, y = -log10(P))) +
  geom_point(alpha = 0.6, size = 1) +
  geom_hline(yintercept = -log10(sig_line), linetype = "dashed", color = "firebrick") +
  labs(title = "Chromosome 5 (expected: nothing here)", x = "Position") +
  theme_minimal(base_size = 12)
```

## Discussion — bring these to the group

**🔧 YOUR TURN**, no code required, just answers:

1. Why might a SNP with a real biological effect on disease risk still
   fail to cross the genome-wide significance line in a study this size?
   Name at least two contributing factors.
2. The significance threshold used throughout this workshop
   ($\approx 2\times10^{-6}$) is far less strict than the standard
   $5\times10^{-8}$ used in most published human GWAS. Why is that
   difference justified here, and why would using $2\times10^{-6}$ on a
   real ~1-million-SNP genotyping array be a problem?
3. Real common-disease GWAS often need tens or hundreds of thousands of
   participants. Point to something specific in today's results that
   illustrates why sample size matters this much.
4. In your own words: what did removing ancestry outliers actually fix in
   notebook 04, mechanically? (Not just "it removed the false positive" —
   *why* did that false positive exist in the first place?)

_Your answers:_ ___FILL_IN___

## Wrap-up

You've now run a (small, simulated) GWAS start to finish: sample QC,
ancestry-based filtering with a reference panel, variant-level MAF/HWE
filtering, association testing, and result interpretation — including
seeing, first-hand, how population stratification creates false positives
and how correcting for it changes both a single SNP's p-value and the
genome-wide inflation statistic. That pipeline, at a much larger scale, is
essentially what underlies every published GWAS you'll read.

When you're ready, compare your thresholds and findings against
`SOLUTIONS.md` in the repo root.
