---
title: "05 — Interpreting the GWAS Hits"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# The setup chunk runs in this file's own directory; every *later*
# chunk runs from the project root thanks to root.dir below.
proj_root <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj_root)
library(dplyr)
library(ggplot2)
source(file.path(proj_root, "R/plot_utils.R"))
```

We'll work only with the **ancestry-corrected** results from notebook 04
(`results/gwas_corrected`) from here on — this is the analysis you'd
actually stand behind.

```{r}
corrected <- read_glm(Sys.glob("results/gwas_corrected.*.glm.logistic*")[1])
sig_line <- 1e-6
```

## Which SNPs reached significance?

```{r}
hits <- corrected %>%
  filter(P < sig_line) %>%
  arrange(P) %>%
  select(ID, CHROM, POS, A1, OR, P)
hits
```

For each hit, sanity-check the reported odds ratio: **which allele is it
relative to?** The `A1` column tells you which allele PLINK tested. If an
`OR` comes out less than 1, that means the *other* allele (not `A1`) is
the one associated with higher risk — take the reciprocal (`1/OR`) to
express the effect in terms of the risk-increasing allele. This isn't a
bug in the simulation; it's a real and common gotcha when reading any
GWAS summary statistics table, since which allele ends up labeled A1 is
just a bookkeeping choice made upstream.

```{r}
hits <- hits %>%
  mutate(OR_risk_allele = ifelse(OR < 1, 1 / OR, OR))
hits
```

## How many chromosomes show a real peak?

```{r, fig.height=4}
plot_manhattan(corrected, title = "Corrected GWAS", sig_line = sig_line)
```

**🔧 YOUR TURN:** count how many *distinct* chromosomes have at least one
SNP crossing the significance line:

```{r}
hits %>% distinct(CHROM)
```

Recall from notebook 00 that this dataset embeds SNPs across a **range**
of allele frequencies and effect sizes, deliberately chosen so that not
every real signal would be expected to reach significance at this sample
size. You should find **three** chromosomes with a peak crossing the line.
There are more than three real causal variants in this data — the ones you
*don't* see are not a failure of your analysis, they're underpowered. That
gap between "how many real effects exist" and "how many you can detect" is
the single most important practical fact about GWAS, and notebook 05's
discussion questions come back to it.

## A closer look: zoom into one peak

```{r}
top_chr <- hits$CHROM[1]

corrected %>%
  filter(CHROM == top_chr) %>%
  ggplot(aes(x = POS, y = -log10(P))) +
  geom_point(alpha = 0.6, size = 1) +
  geom_hline(yintercept = -log10(sig_line), linetype = "dashed", color = "firebrick") +
  labs(title = paste("Chromosome", top_chr, "region"), x = "Position") +
  theme_minimal(base_size = 12)
```

Notice that the peak here is a **single isolated point** — its immediate
neighbours sit right down in the noise. That is a deliberate property of
this simulation, and it's worth pausing on, because it is *not* what real
GWAS data looks like.

Every SNP in this dataset was simulated independently, so there is no
**linkage disequilibrium (LD)** — no correlation between neighbouring
markers. In real data, a causal variant drags its neighbours up with it:
SNPs physically close by are inherited together, so they carry a partial
copy of the same signal and form the characteristic "skyscraper with
shoulders" shape you see in published Manhattan plots. That's also why
real GWAS hits are reported as *loci* rather than single variants, and why
the top SNP at a locus is very often not the causal one — it's just the
best-tagged marker nearby.

Two things in this workshop are quietly shaped by that absence: the
LD-pruning step in notebook 02 had almost nothing to remove, and our
significance threshold treats all ~50,000 SNPs as independent tests, which
here is actually true (in real data it's conservative, because correlated
SNPs aren't independent tests).

Compare the single-point peak to what a chromosome with *no* real signal
looks like:

```{r}
corrected %>%
  filter(CHROM == 10) %>%   # chromosome 10 was built with no injected signal at all
  ggplot(aes(x = POS, y = -log10(P))) +
  geom_point(alpha = 0.6, size = 1) +
  geom_hline(yintercept = -log10(sig_line), linetype = "dashed", color = "firebrick") +
  labs(title = "Chromosome 10 (expected: nothing here)", x = "Position") +
  theme_minimal(base_size = 12)
```

## Discussion — bring these to the group

**🔧 YOUR TURN**, no code required, just answers:

1. Why might a SNP with a real biological effect on disease risk still
   fail to cross the genome-wide significance line in a study this size?
   Name at least two contributing factors.
2. The significance threshold used throughout this workshop
   ($\approx 1\times10^{-6}$) is far less strict than the standard
   $5\times10^{-8}$ used in most published human GWAS. Why is that
   difference justified here, and why would using $1\times10^{-6}$ on a
   real ~1-million-SNP genotyping array be a problem?
3. Real common-disease GWAS often need tens or hundreds of thousands of
   participants. Point to something specific in today's results that
   illustrates why sample size matters this much.
4. In your own words: what did removing ancestry outliers actually fix in
   notebook 04, mechanically? (Not just "it removed the false positive" —
   *why* did that false positive exist in the first place?)

_Your answers:_ ___FILL_IN___

## Wrap-up

You've now run a (small, simulated) GWAS start to finish: sample QC,
ancestry-based filtering with a reference panel, variant-level MAF/HWE
filtering, association testing, and result interpretation — including
seeing, first-hand, how population stratification creates false positives
and how correcting for it changes both a single SNP's p-value and the
genome-wide inflation statistic. That pipeline, at a much larger scale, is
essentially what underlies every published GWAS you'll read.

When you're ready, compare your thresholds and findings against
`SOLUTIONS.md` in the repo root.
