---
title: "05 — Interpreting the GWAS Hits"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Locate the project root and work from there. This makes the notebook behave
# the same whether you press Knit or run chunks one at a time, and whether
# RStudio's working directory is the project root or this notebooks/ folder.
proj_root <- if (file.exists("R/sim_utils.R")) normalizePath(".") else normalizePath("..")
if (!file.exists(file.path(proj_root, "R", "sim_utils.R"))) {
  stop("Can't find the project root from ", getwd(), ".\n",
       "Fix: in the Console, run  setwd(\"/workspaces/popgen-gwas-tutorial\")  ",
       "then run this chunk again. (Or: click popgen-gwas-tutorial.Rproj in the ",
       "Files pane to reopen as a project, which sets this automatically from now on.)")
}
knitr::opts_knit$set(root.dir = proj_root)                     # applies when knitting
if (!isTRUE(getOption("knitr.in.progress"))) setwd(proj_root)  # applies when running chunks
library(dplyr)
library(ggplot2)
source(file.path(proj_root, "R/plot_utils.R"))
```

We'll work only with the **ancestry-corrected** results from notebook 04
(`results/gwas_corrected`) from here on — this is the analysis you'd
actually stand behind.

```{r}
gwas_file <- Sys.glob("results/gwas_corrected.*.glm.logistic*")[1]
if (is.na(gwas_file)) {
  stop("No corrected GWAS results found in results/. ",
       "Run notebook 04 (04_association_gwas.Rmd) first — this notebook ",
       "reads the output it produces.")
}
corrected <- read_glm(gwas_file)
sig_line <- 2e-6
```

## Which SNPs reached significance?

```{r}
hits <- corrected %>%
  filter(P < sig_line) %>%
  arrange(P) %>%
  select(ID, CHROM, POS, A1, OR, P)
hits
```

```{r}
if (nrow(hits) == 0) {
  warning("No SNPs crossed your significance threshold (", sig_line, "). ",
          "The plots below need at least one hit. Try a less strict threshold, ",
          "or check that notebook 04's corrected analysis ran on data/study_final.")
}
```

For each hit, sanity-check the reported odds ratio: **which allele is it
relative to?** The `A1` column tells you which allele PLINK tested. If an
`OR` comes out less than 1, that means the *other* allele (not `A1`) is
the one associated with higher risk — take the reciprocal (`1/OR`) to
express the effect in terms of the risk-increasing allele. (On this
dataset's seed, every current hit happens to already have `OR > 1`, so the
line below is a no-op here — but don't skip it: which allele lands as A1
is an upstream bookkeeping choice, not something you control, and it
won't always land this conveniently. This is a real and common gotcha when
reading any GWAS summary statistics table.)

```{r}
hits <- hits %>%
  mutate(OR_risk_allele = ifelse(OR < 1, 1 / OR, OR))
hits
```

## How many chromosomes show a real peak?

```{r, fig.height=4}
plot_manhattan(corrected, title = "Corrected GWAS", sig_line = sig_line)
```

**🔧 YOUR TURN:** count how many *distinct* SNPs, and how many *distinct*
chromosomes, have at least one hit crossing the significance line:

```{r}
hits %>% distinct(CHROM)
nrow(hits)
```

Recall from notebook 00 that this dataset embeds SNPs across a **range**
of allele frequencies and effect sizes, deliberately chosen so that not
every real signal would be expected to reach significance at this sample
size. You should find **5 significant SNPs across 2 chromosomes** — notice
those two numbers aren't the same, for two different reasons you'll see up
close below: (1) more than one real causal SNP can share a chromosome by
design (chromosome 2 has two here), and (2) a SNP *right next to* a real
causal SNP can ride along and cross the line too, without being causal
itself. Counting distinct *SNPs* is what actually matters for "how many
signals did I find"; counting distinct chromosomes undercounts whenever
either of those happens. And there are more than 5 real causal variants in
this data in the first place — the ones you don't see at all are not a
failure of your analysis, they're underpowered. That gap between "how many
real effects exist" and "how many you can detect" is the single most
important practical fact about GWAS, and this notebook's discussion
questions come back to it.

## A closer look: zoom into the top chromosome

```{r}
top_chr <- hits$CHROM[1]

corrected %>%
  filter(CHROM == top_chr) %>%
  ggplot(aes(x = POS, y = -log10(P))) +
  geom_point(alpha = 0.6, size = 1) +
  geom_hline(yintercept = -log10(sig_line), linetype = "dashed", color = "firebrick") +
  labs(title = paste("Chromosome", top_chr, "region"), x = "Position") +
  theme_minimal(base_size = 12)
```

Two separate peaks on this one chromosome — two different causal SNPs
happen to have been placed here. At this zoomed-out scale they look like
thin spikes, but look closer and each one is actually a small **hill**,
not a needle. Zoom into just the tallest one:

```{r}
top_snp <- hits$POS[1]

corrected %>%
  filter(CHROM == top_chr, POS > top_snp - 30000, POS < top_snp + 30000) %>%
  ggplot(aes(x = POS, y = -log10(P))) +
  geom_point(alpha = 0.7, size = 1.5) +
  geom_hline(yintercept = -log10(sig_line), linetype = "dashed", color = "firebrick") +
  labs(title = "Zoomed in: the top hit's immediate neighbourhood", x = "Position") +
  theme_minimal(base_size = 12)
```

That's the "skyscraper with shoulders" shape real GWAS Manhattan plots are
known for: nearby SNPs build up toward the peak and fall back off on the
other side, instead of the causal SNP standing alone against flat noise.
This is **linkage disequilibrium (LD)**: SNPs physically close together
tend to be inherited as a block rather than independently, so they carry a
correlated, diluted copy of the same signal. Real GWAS hits are reported
as *loci* rather than single variants for exactly this reason — LD makes
it genuinely ambiguous, from the association test alone, which SNP in a
correlated neighbourhood is the actual causal one, and pinning that down
(*fine-mapping*) is its own hard problem beyond just finding the locus.

This dataset now has that property, but only in **small, deliberate
patches around the causal SNPs** — everywhere else, SNPs still behave
completely independently (unlike real genomes, where LD extends
genome-wide in a complex block structure). That's why LD-pruning in
notebook 02 only removed a small number of SNPs, clustered right around
these same peaks, rather than the large fraction it would strip from real
data.

**🔧 YOUR TURN:** look at the two SNPs highlighted below and compare their
p-values:

```{r}
corrected %>% filter(ID %in% c("chr1_snp02500", "chr1_snp02501")) %>%
  select(ID, CHROM, POS, P)
```

One of these is the deliberately-placed causal SNP; the other is its
immediate neighbour, one position over, tagging along via LD closely
enough to also cross the significance line. In this particular case the
true causal SNP is still the more significant of the two — but that's not
guaranteed in general. In real data, a neighbour can sometimes show a
*stronger* signal than the actual causal variant (due to how well each SNP
happens to be genotyped or tagged), which is exactly why "the top hit at a
locus" and "the causal variant at a locus" are different claims, and why
GWAS papers report *loci*, then spend separate effort fine-mapping which
specific variant within each locus is doing the actual biological work.

Compare all of this — real shoulders, a locus with more than one
significant SNP — to what a chromosome with *no* injected signal at all
looks like:

```{r}
corrected %>%
  filter(CHROM == 5) %>%   # chromosome 5 was built with no injected signal at all
  ggplot(aes(x = POS, y = -log10(P))) +
  geom_point(alpha = 0.6, size = 1) +
  geom_hline(yintercept = -log10(sig_line), linetype = "dashed", color = "firebrick") +
  labs(title = "Chromosome 5 (expected: nothing here)", x = "Position") +
  theme_minimal(base_size = 12)
```

## Discussion — bring these to the group

**🔧 YOUR TURN**, no code required, just answers:

1. Why might a SNP with a real biological effect on disease risk still
   fail to cross the genome-wide significance line in a study this size?
   Name at least two contributing factors.
2. The significance threshold used throughout this workshop
   ($\approx 2\times10^{-6}$) is far less strict than the standard
   $5\times10^{-8}$ used in most published human GWAS. Why is that
   difference justified here, and why would using $2\times10^{-6}$ on a
   real ~1-million-SNP genotyping array be a problem?
3. Real common-disease GWAS often need tens or hundreds of thousands of
   participants. Point to something specific in today's results that
   illustrates why sample size matters this much.
4. In your own words: what did removing ancestry outliers actually fix in
   notebook 04, mechanically? (Not just "it removed the false positive" —
   *why* did that false positive exist in the first place?)

_Your answers:_ ___FILL_IN___

## Wrap-up

You've now run a (small, simulated) GWAS start to finish: sample QC,
ancestry-based filtering with a reference panel, variant-level MAF/HWE
filtering, association testing, and result interpretation — including
seeing, first-hand, how population stratification creates false positives
and how correcting for it changes both a single SNP's p-value and the
genome-wide inflation statistic. That pipeline, at a much larger scale, is
essentially what underlies every published GWAS you'll read.

When you're ready, compare your thresholds and findings against
`SOLUTIONS.md` in the repo root.
